---
title: "DESeq2_lung"
output:
  html_notebook:
    toc: yes
    theme: united
---
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
This is the analysis of the lung data set provided by the authors of MLSeq using DESeq2.



Setup

```{r, eval = FALSE}
## ----setup, echo=FALSE, results="hide"----------------------------------------
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE,
                      error = FALSE,
                      warning = TRUE)
```
```{r}
library("DESeq2")
library("vsn")
library("hexbin")
library("pheatmap")
library("RColorBrewer")
library("ggplot2")
```
Making it possible to run both on local machine as well as on cluster
```{r}
locale <- FALSE
recompute <- 1

prefix <- ifelse(locale, "~", "")
dir <- file.path(prefix, "data/bioinf/projects/data/2022_MLseq_KP/lung_DESeq2")
data_dir <- file.path(dir, "data/")
results_dir <- file.path(dir, "results/")
graphics_dir <- file.path(dir, "graphics/")
```
Load and read file
The lung data set is a count matrix, to use it in DESeq2 analysis we use the method "DESeqDataSetFromMatrix" which expects the following inputs:
A) The count matrix
B) A data.frame/ DataFrame object
C) A design argument
Since there is no data.frame object provided for the lung data, we need to provide it. To do so, we create a table containing one column for the sample number and one column for the treatment.
```{r}
recompute <- 0
file <- file.path(data_dir, "lung.csv")
file_lung <- file.path(data_dir, "01_lung.Rdata")
if(recompute){
  lung <- read.csv(file,  sep=" ")
  save(lung, file = file_lung)
} else {
  load(file_lung)
}
recompute <- 1
#the first 29 genes of the dataset are unannotated so I remove them
lung <- lung[30:20531, ]

#I furthermore remove the numbers in the gene names
name <- sub("[|].*", "", (rownames(lung)))
rownames(lung) <- make.names(name, unique = TRUE)
#creating the data.frame object
#lung data has 576 samples in T1 and 552 in T2.
class <- data.frame(condition = factor(rep(c("T1","T2"), c(576, 552))))
colnames(lung) <- 1:1128 #replacing name of columns so they match data.frame

#read the data into a DESeqDataSet
file_dds <- file.path(data_dir, "02_dds.Rdata")
if (recompute){
  dds <- DESeqDataSetFromMatrix(countData = lung,
                              colData = class,
                              design = ~ condition)
  save(dds, file = file_dds)
} else {
  load(file_dds)
}

#look at object
dds

```
Pre-filtering
by excluding low count genes we reduce the memory usage of the later methods
```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```
This reduced genes from 20500 to 20100

set factor levels
It may be important to define in which order the conditions are compared. (default is alphabetical order)
```{r}
dds$condition <- factor(dds$condition, levels = c("T1","T2"))
```
The following steps are to visualize and cluster the data. For this process a transformed version of the data is useful.
5. rlog-transformation
Quote from DESeq2 paper: we use the shrinkage approach of DESeq2 to implement a regularized logarithm transformation (rlog), which behaves similarly to a log2 transformation for genes with high counts, while shrinking together the values for different samples for genes with low counts. It therefore avoids a commonly observed property of the standard logarithm transformation, the spreading apart of data for genes with low counts, where random noise is likely to dominate any biologically meaningful signal.
(For a big number of samples there is an alternative: VST ~  variance stabilizing transformation)
```{r}
recompute <- 1
file_vsd <- file.path(results_dir, "03_vsd.Rdata")
file_rld <- file.path(data_dir, "04_rld.Rdata")
if (recompute) {
  vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
   #rld takes incredibly long for this dataset, so I'm using the vst instead here
  x <- Sys.time()
  rld <- rlog(dds, blind = FALSE)
  time_rld <- Sys.time()-x
  print(time_rld)
  save(vsd, file = file_vsd)
  save(rld, file = file_rld)
} else {
  load(file = file_vsd)
  #load(file = file_rld)
}
head(assay(rld), 3)

```
6. Plot showing dependence of sd on mean
```{r}
#plot of sd and mean
ntd <- normTransform(dds)


jpeg(file = file.path(graphics_dir, "02_rld_meanSd.jpeg"))
meanSdPlot(assay(rld))
dev.off()
jpeg(file = file.path(graphics_dir, "03_vsd_meanSd.jpeg"))
meanSdPlot(assay(vsd))
dev.off()
```
vsd has the most even mean (could not do rld)

8. Heatmap of the sample-to-sample distances
sampleDists gives distances
creating plot with RColorBrewer and pheatmap

```{r lung s2s}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)

rownames(sampleDistMatrix) <- paste(vsd$condition)
colnames(sampleDistMatrix) <- colnames(vsd)
annotation <- data.frame(condition =class$condition)
save(annotation, file = file.path(results_dir, "annotation"))
#rownames(annotation) <- rownames(sampleDistMatrix)
colors <- colorRampPalette( rev(brewer.pal(9, "Greys")) )(255)
# save heatmap in image file
file <- paste0(graphics_dir, "04_heatmap_s2s-distance.pdf")

pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         treeheight_row = 0,
         treeheight_col = 20,
         legend = T,
         cellwidth=10,
         cellheight=10,
         height = 1000,
         width = 1000,
         #annotation_row = annotation,
         annotation_col = annotation,
         #annotation_colors = ann_colors,
         show_colnames = T,
         show_rownames = F,
         #main=paste0("Sample-2-sample distance vsd-transformed\n"),
         #filename=file,
         color=colors
)

```

PCA
Shows samples in 2D plot, the axes representing their two principal components
```{r}

jpeg(file=file.path(graphics_dir, "06_vsd_PCA.jpeg"))
plotPCA(vsd)# + geom_text(aes(label = name), vjust = 2)
dev.off()
```
 Clustering is completed.
Save dds for MLSeq analysis
```{r}
save(dds, file = file.path(results_dir, "03_dds_pre_processed.Rdata"))
```

Standard differential expression analysis
This method goes through all steps of DESeq2 and gives results.
```{r}


recompute <- 0
file_res <- file.path(results_dir, "03_res.Rdata")
file_dds <- file.path(results_dir, "04_dds.Rdata")
if (recompute) {
  start <- Sys.time()
  dds <- DESeq(dds)
  res <- results(dds)
  save(dds, file = file_dds)
  save(res, file = file_res)
} else {
  load(file = file_dds)
  load(file = file_res)
}

res
```
12. Perform LFC transformation
For better visualization and gene ranking we use the method lfcShrink
```{r}
resultsNames(dds)
resLFC <- lfcShrink(dds, coef = "condition_T2_vs_T1", type = "apeglm")
resLFC
```

13. Table ordered by p-value
For a better overview we can order the table by p-value and give a summary of how many p-values were <0.1
Finally, we can get a results table.
```{r}
resOrdered <- res[order(res$pvalue),]
resOrdered
summary(res)
#How many below 0.1?
sum(res$padj < 0.1, na.rm = TRUE)
#results table
res05 <- results(dds, alpha = 0.01)
summary(res05)
#sum(res05$padj < 0.05, na.rm = TRUE)
```
16. Export the list of genes that pass p-value (adjusted with Benjamini-Hochberg procedure)
```{r}
resSig <- subset(resOrdered, padj < 0.1)
resSig
write.csv(as.data.frame(resSig), file = file.path(results_dir, "03_results_lung_p01"))
write.csv(as.data.frame(resOrdered), file = file.path(results_dir, "07_results_all.csv"))
#runtime
if(recompute){
  runtime <- Sys.time() - start
}
capture.output(runtime, file = file.path(results_dir, "08_runtime.txt"))
print("done")
```