---
title: "DESeq2_cervical
output: html_DESeq2_cervical
---
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
This is the analysis of the cervical data set provided by the authors of MLSeq using DESeq2.



1. load libraries

```{r}
## ----setup, echo=FALSE, results="hide"----------------------------------------
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
library("DESeq2")
```
2. load and read file
The cervical data set is a count matrix, to use it in DESeq2 analysis we use the method "DESeqDataSetFromMatrix" which expects the following inputs:
A) The count matrix
B) A data.frame/ DataFrame object
C) A design argument that I don't fully understand but which seems to be necessary
Since there is no data.frame object provided for the cervical data, we need to provide it. To do so, we create a table containing one column for the sample number and one column for the treatment. The first 29 samples are controls, the last 29 are treated.
```{r}
load(file="/home/pilz/data/MLSeq_original_data/cervical.rda")
#creating the data.frame object
class <- data.frame(condition = factor(rep(c("untreated","treated"), c(29, 29))))
#colnames(cervical) <- 1:58 #replacing name of columns so they match data.frame
#optional: check data.frame object and beginning of the data
class
head(cervical)
#read the data into a DESeqDataSet
dds <- DESeqDataSetFromMatrix(countData = cervical,
                              colData = class,
                              design = ~ condition)
#look at object
dds

```
3. Pre-filtering
by excluding low count genes we reduce the memory usage of the later methods
```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```
4. set factor levels
It may be important to make sure in which order the conditions are compared. (default is alphabetical order)
```{r}
dds$condition <- factor(dds$condition, levels = c("untreated","treated"))
```
5. Standard differential expression analysis
This method goes through all steps of DESeq2 and gives results.
```{r}
dds <- DESeq(dds)
res <- results(dds)
res
```
6. Perform LFC transformation
For better visualization and gene ranking we use the method lfcShrink
```{r}
resultsNames(dds)
resLFC <- lfcShrink(dds, coef="condition_treated_vs_untreated", type="apeglm")
resLFC
```
7. Table ordered by p-value
For a better overview we can order the table by p-value and give a summary of how many p-values were <0.1
Finally, we can get a results table.
```{r}
resOrdered <- res[order(res$pvalue),]
resOrdered
summary(res)
#How many below 0.1?
sum(res$padj < 0.1, na.rm=TRUE)
#results table
res05 <- results(dds, alpha=0.05)
summary(res05)
sum(res05$padj < 0.05, na.rm=TRUE)
```
After this simple analysis, we will now go on to take the steps in between individually.
8. rlog-transformation
Apparently, you do this if you don't want to do DE but visualization and clustering instead. Main point is to remove the high variance for low means
rlog ~ regularized logarithm - normalizes on basis of library size and transforms data to log2 scale
(For a big number of samples there is an alternative: VST ~  variance stabilizing transformations)
```{r}
rld <- rlog(dds, blind=FALSE)
head(assay(rld), 3)

```
9. Plot showing dependence of sd on mean
```{r}
#plot of sd and mean
library("vsn")
meanSdPlot(assay(rld))
```
This is not what I expected. The sd should be normalized -> about equal for any mean
10. Heatmap
```{r}
library("pheatmap")
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("condition")])
pheatmap(assay(rld)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
```
11. Heatmap of the sample-to-sample distances
sampleDists gives distances
creating plot with RColorBrewer and pheatmap
```{r}
sampleDists <- dist(t(assay(rld)))
#plot
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(rld$condition)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

```
12. PCA
Shows samples in 2D plot, the axes representing their two principal components
```{r}
plotPCA(rld, intgroup=c("condition"))
```
13. Exlude samples
14. DESeq()
Soll ich die Analyse jetzt mit den normalisierten Daten oder dem Original ohne excludierte samples ausführen?
15. Standard differential expression analysis
This method goes through all steps of DESeq2 and gives results.
Let's try with rld - does not work: rld is no DESeq2 data set
redo steps 5 to 7
- 5.
```{r}
rld <- DESeq(rld)
res <- results(rld)
res
```
6. Perform LFC transformation
For better visualization and gene ranking we use the method lfcShrink
```{r}
resultsNames(dds)
resLFC <- lfcShrink(dds, coef="condition_treated_vs_untreated", type="apeglm")
resLFC
```
7. Table ordered by p-value
For a better overview we can order the table by p-value and give a summary of how many p-values were <0.1
Finally, we can get a results table.
```{r}
resOrdered <- res[order(res$pvalue),]
resOrdered
summary(res)
#How many below 0.1?
sum(res$padj < 0.1, na.rm=TRUE)
#results table
res05 <- results(dds, alpha=0.05)
summary(res05)
sum(res05$padj < 0.05, na.rm=TRUE)
```
