---
title: "DESeq2_cervical"
output:
  html_notebook:
    toc: yes
    theme: united
---
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
This is the analysis of the cervical data set provided by the authors of MLSeq using DESeq2.



1. setup

```{r, eval = FALSE}
## ----setup, echo=FALSE, results="hide"----------------------------------------
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE,
                      error = FALSE,
                      warning = TRUE)
```
```{r}
library("DESeq2")
library("vsn")
library("hexbin")
library("pheatmap")
library("RColorBrewer")
library("ggplot2")
```
Making it possible to run both on local machine and on cluster
```{r}
local <- FALSE
recompute <- 1

prefix <- ifelse(local, "~", "")
dir <- file.path(prefix, "data/bioinf/projects/data/2022_MLseq_KP/cervical_DESeq2")
data_dir <- file.path(dir, "data/")
results_dir <- file.path(dir, "results/")
graphics_dir <- file.path(dir, "graphics/")
```
2. load and read file
The cervical data set is a count matrix, to use it in DESeq2 analysis we use the method "DESeqDataSetFromMatrix" which expects the following inputs:
A) The count matrix
B) A data.frame/ DataFrame object
C) A design argument that I don't fully understand. It is used to estimate dispersion ranges
Since there is no data.frame object provided for the cervical data, we need to provide it. To do so, we create a table containing one column for the sample number and one column for the treatment. The first 29 samples are controls, the last 29 are treated.
```{r}
file <- file.path(data_dir, "cervical.rda")
load(file = file)
#creating the data.frame object
class <- data.frame(condition = factor(rep(c("untreated", "treated"), c(29, 29))))
colnames(cervical) <- 1:58 #replacing name of columns so they match data.frame
#optional: check data.frame object and beginning of the data
class
head(cervical)
#read the data into a DESeqDataSet
dds <- DESeqDataSetFromMatrix(countData = cervical,
                              colData = class,
                              design = ~condition)
#look at object
dds

```
3. Pre-filtering
By excluding low count genes we reduce the memory usage and compute requirement of the later methods
Note that filtering is applied when using the DESeq() function to increase statistical power
```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```
4. set factor levels
It may be important to define in which order the conditions are compared. (default is alphabetical order)
```{r}
dds$condition <- factor(dds$condition, levels = c("untreated", "treated"))
```
The following steps are to visualize and cluster the data. For this process a transformed version of the data is useful.
5. rlog-transformation
You do this if you don't want to do DE but visualization and clustering instead. Main point is to remove the high variance for low means
rlog ~ regularized logarithm - normalizes on basis of library size and transforms data to log2 scale
Quote from DESeq2 paper: we use the shrinkage approach of DESeq2 to implement a regularized logarithm transformation (rlog), which behaves similarly to a log2 transformation for genes with high counts, while shrinking together the values for different samples for genes with low counts. It therefore avoids a commonly observed property of the standard logarithm transformation, the spreading apart of data for genes with low counts, where random noise is likely to dominate any biologically meaningful signal.
(For a big number of samples there is an alternative: VST ~  variance stabilizing transformation)
```{r}
file_vsd <- file.path(results_dir, "01_vsd.Rdata")
file_rld <- file.path(results_dir, "02_rld.Rdata")
if (recompute) {
  vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
  rld <- rlog(dds, blind = FALSE)
  save(vsd, file = file_vsd)
  save(rld, file = file_rld)
} else {
  load(file = file_vsd)
  load(file = file_rld)
}
head(assay(rld), 3)

```
6. Plot showing dependence of sd on mean
```{r}
#plot of sd and mean
ntd <- normTransform(dds)

#meanSdPlot(assay(ntd))
jpeg(file = file.path(graphics_dir, "01_ntd_meanSd.jpeg"))
meanSdPlot(assay(ntd))
dev.off()
jpeg(file = file.path(graphics_dir, "02_rld_meanSd.jpeg"))
meanSdPlot(assay(rld))
dev.off()
jpeg(file = file.path(graphics_dir, "03_vsd_meanSd.jpeg"))
meanSdPlot(assay(vsd))
dev.off()
```
rld has the most even mean
7. Heatmap
We need to first estimate the size factors to make it work
```{r}

dds <- estimateSizeFactors(dds)
select <- order(rowMeans(counts(dds, normalized = TRUE)),
                decreasing = TRUE)[1:20]
df <- as.data.frame(colData(dds)[, c("condition")])
#jpeg(file="/home/pilz/data/graphics/heatmap.jpeg")
#saving the pheatmap is complex
heatmap <- pheatmap(assay(rld)[select,],
                    cluster_rows = TRUE,
                    show_rownames = FALSE,
                    cluster_cols = TRUE,
                    annotation_col = class)
save_pheatmap_pdf <- function(x, filename, width = 7, height = 7) {
  stopifnot(!missing(x))
  stopifnot(!missing(filename))
  pdf(filename, width = width, height = height)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}
save_pheatmap_pdf(heatmap, file.path(graphics_dir, "04_rld_heatmap.pdf"))
#dev.off()
```
Observation: Sample 9 looks off. Samples 35 and 39 do too (although only slightly)
8. Heatmap of the sample-to-sample distances
sampleDists gives distances
creating plot with RColorBrewer and pheatmap
```{r}
sampleDists <- dist(t(assay(rld)))
#plot

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(rld$condition)#paste(rld$colnames)
colnames(sampleDistMatrix) <- colnames(rld)
annotation <- data.frame(condition = class$condition)
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
heatmap2 <- pheatmap(sampleDistMatrix,
                     clustering_distance_rows = sampleDists,
                     clustering_distance_cols = sampleDists,
                     #treeheight_row = 20,
                     #treeheight_col = 20,
                     #legend = T,
                     #cellwidth=10,
                     #cellheight=10,
                     #annotation_row = class,
                     annotation_col = annotation,
                     #annotation_colors = ann_colors,
                     show_colnames = T,
                     show_rownames = F,
                     col = colors)
save_pheatmap_pdf(heatmap2, file.path(graphics_dir, "05_rld_heatmap_sampledist.pdf"))
```
Observation: Sample 54, 15 and 29 look out of place
9. PCA
Shows samples in 2D plot, the axes representing their two principal components
```{r}

plotPCA(rld) + geom_text(aes(label = name), vjust = 2)
jpeg(file=file.path(graphics_dir, "06_rld_PCA.jpeg"))
plotPCA(rld) + geom_text(aes(label = name), vjust = 2)
dev.off()
```
Observation: Sample 15 is a clear outlier on the y axis As it could also be observed as an outlier in the sample-to-sample distance, it will be excluded.
Exclude sample 15 and redo PCA
```{r}
#maybe an alternative:
#keep <- rowSums(counts(dds)) >= 10
#dds <- dds[keep,]

rld <- rld[, -15]
dds <- dds[, -15]

plotPCA(rld) + geom_text(aes(label = name), vjust = 2)
jpeg(file = file.path(graphics_dir, "07_rld_PCA.jpeg"))
plotPCA(rld) + geom_text(aes(label = name), vjust = 2)
dev.off()
```
Observation: Sample 54 is a clear outlier, it also was observed to be different from the others in the sts heatmap
```{r}
rld <- rld[, -53] #since a sample before it was excluded, 54 is in position 53
dds <- dds[, -53]

plotPCA(rld) + geom_text(aes(label = name), vjust = 2)
jpeg(file = file.path(graphics_dir, "08_rld_PCA.jpeg"))
plotPCA(rld) + geom_text(aes(label = name), vjust = 2)
dev.off()
```
Now, 29 also looks like an outlier. It too was an outlier in the sts distance
```{r}
rld <- rld[, -28] #again, 29 is now at position 28 because 15 was excluded
dds <- dds[, -28]

plotPCA(rld) + geom_text(aes(label = name), vjust = 2)
jpeg(file = file.path(graphics_dir, "09_rld_PCA.jpeg"))
plotPCA(rld) + geom_text(aes(label = name), vjust = 2)
dev.off()
```
No obvious outliers anymore. Clustering is completed.

13. Standard differential expression analysis
This method goes through all steps of DESeq2 and gives results.
```{r}
file_res <- file.path(results_dir, "03_res.Rdata")
file_dds <- file.path(results_dir, "04_dds.Rdata")
if (recompute) {
  dds <- DESeq(dds)
  res <- results(dds)
  save(dds, file = file_dds)
  save(res, file = file_res)
} else {
  load(file = file_dds)
  load(file = file_res)
}

res
```
12. Perform LFC transformation
For better visualization and gene ranking we use the method lfcShrink
```{r}
resultsNames(dds)
resLFC <- lfcShrink(dds, coef = "condition_treated_vs_untreated", type = "apeglm")
resLFC
```

13. Table ordered by p-value
For a better overview we can order the table by p-value and give a summary of how many p-values were <0.1
Finally, we can get a results table.
```{r}
resOrdered <- res[order(res$pvalue),]
resOrdered
summary(res)
#How many below 0.1?
sum(res$padj < 0.1, na.rm = TRUE)
#results table
res05 <- results(dds, alpha = 0.05)
summary(res05)
sum(res05$padj < 0.05, na.rm = TRUE)
```
(optional: Independent hypothesis weighting)
14. MA-plot
```{r}
jpeg(file = file.path(graphics_dir, "10_MA_plot"))
plotMA(res, ylim = c(-10, 10))
dev.off()
```
This looks nothing like what I expected. I looks like it is mirrored by the y-axis and moved -1.6 into x-direction

Same plot for shrunken log2 fold chages (removes noise)
```{r}
jpeg(file = file.path(graphics_dir, "11_MA_plot_weighting_LFC"))
plotMA(resLFC, ylim = c(-10, 10))
dev.off()
```
Does not work due to Error in 12
15. Plot counts for individual genes
```{r}
jpeg(file = file.path(graphics_dir, "12_counts_per_gene"))
plotCounts(dds, gene = which.min(res$padj), intgroup = "condition")
dev.off()
```
16. Export the list of genes that pass p-value
```{r}
resSig <- subset(resOrdered, padj < 0.05)
resSig
write.csv(as.data.frame(resSig), file = file.path(results_dir, "05_results_cervical_p05_no_exclusion.csv"))
write.csv(as.data.frame(resOrdered), file = file.path(results_dir, "06_results_cervical_all.csv"))
```
