---
title: "DESeq2_lung"
output:
  html_notebook:
    toc: yes
    theme: united
---
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
This is the analysis of the MAIT project using DESeq2.


1. setup

```{r, eval = FALSE}
## ----setup, echo=FALSE, results="hide"----------------------------------------
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = TRUE,
                      error = TRUE,
                      echo = TRUE,
                      warning = TRUE)
```
```{r}
library("DESeq2")
library("vsn")
library("hexbin")
library("pheatmap")
library("RColorBrewer")
#library("ggplot2")
library("tidyverse")
```
Making it possible to run both on local machine as well as on cluster
```{r}
locale <- FALSE
recompute <- 1

prefix <- ifelse(locale, "~", "")
dir <- file.path(prefix, "data/bioinf/projects/data/2022_MLseq_KP/MAIT_DESeq2")
data_dir <- file.path(dir, "data/")
results_dir <- file.path(dir, "results/")
graphics_dir <- file.path(dir, "graphics/")
```
Load and read file
Roughly following the script https://git.ufz.de/rnaseq/2018_maitcellsrnaseq/-/blob/main/DE_Analysis/01_DESeq2.Rmd
```{r}
count_files <- list.files(path=data_dir, pattern=".txt", full.names=TRUE)#all files

count_files_0h <- list.files(path=data_dir, pattern="0h", full.names=TRUE)#first timepoint which will be excluded from analysis
count_files_K12 <- list.files(path=data_dir, pattern="K12", full.names=TRUE)#K12 treatment
count_files_CD3 <- list.files(path=data_dir, pattern="CD3", full.names=TRUE)#CD3 treatment
count_files_K12vsUnstim <- count_files [! count_files %in% count_files_0h]
count_files_K12vsUnstim <- count_files_K12vsUnstim [! count_files_K12vsUnstim %in% count_files_CD3]

names(count_files) <- gsub("-htseq_counts.txt", "", basename(count_files)) # this shortens the names
names(count_files_K12vsUnstim) <- gsub("-htseq_counts.txt", "", basename(count_files_K12vsUnstim))
SampleTableFile <- paste0(data_dir, "SampleTable_MAITseq_All.csv")
backgroundInfo <- read.csv(SampleTableFile,  sep=",") #info about gender, age etc



file_sampleTable <- file.path(results_dir, "01_sampleTable")

if(recompute) {
    STcolsOfInterest <- c(1:16)
    STcolsAsFactors <- c(4,8,9,12,16)

    sT <- read.table(SampleTableFile, header=T, sep=",", stringsAsFactors = FALSE)
    for(i in STcolsAsFactors)
    { #select requested columns
      sT[,i] <- as.factor(sT[,i])
    }
    sT <- sT[,STcolsOfInterest]
    sampleTable <- sT
    colnames(sampleTable)[c(2,4,5)] <- c("sampleName", "time", "condition")
    rownames(sampleTable) <- sampleTable$sampleName
    sampleTable$time <- as.factor(paste0(sampleTable$time, "h"))
    # fileName is required by DESeq2 function
    rows <- match(names(count_files), rownames(sampleTable))
    sampleTable$fileName[rows] <- basename(count_files)
    # formatting condition etc.
    sampleTable$condition <- ifelse(sampleTable$condition=='-', "Ctrl_0",
                                    ifelse(sampleTable$condition=='unstim',
                                           "Ctrl_16", sampleTable$condition))
    sampleTable$condition <- gsub("-", "_", sampleTable$condition)
    sampleTable$condition <- as.factor(sampleTable$condition)
    sampleTable$treatment <- ifelse(str_detect(sampleTable$condition,"Ctrl"),
                                    "control", "treated")
    sampleTable$treatment <- as.factor(sampleTable$treatment)
    sampleTable <- sampleTable[,c(2,17,4,5,18,3,6:16)]
    save(sampleTable, file=file_sampleTable)
} else {
    load(file = file_sampleTable)
}
print(paste("Sample table loaded:", SampleTableFile, "-",
            dim(sampleTable)[2], "columns for",
            dim(sampleTable)[1], "samples:"))
sampleTable$condition <- gsub("CD3_28", "CD3-28",
                              gsub("Ctrl_16", "16h_Ctrl",
                                   gsub("Ctrl_0", "0h_Ctrl", sampleTable$condition)))
names(sampleTable)

sampleTable_K12_unstim <- filter(sampleTable, condition=="K12" | condition=="16h_Ctrl")#table for a subset -> only K12 vs unstim


```
```{r}
recompute <- 1
file_dds_K12 <- file.path(data_dir, "02_dds_K12")
if(recompute){

  design <- formula(~ condition)
  dds_K12_raw <- DESeqDataSetFromHTSeqCount(sampleTable=sampleTable_K12_unstim,
                                  directory= data_dir,
                                  design=design)
  save (dds_K12_raw, file= file_dds_K12)


} else {
  load(file_dds_K12)
}

#look at object
dds_K12_raw



```
exclude XLOC
```{r exclude XLOC}
count <- assay(dds_K12_raw)
sub <- count[-grep("XLO", rownames(count)), , drop = FALSE]

#add gene names
rownames(sub) <- sub("[.].*", "", rownames(sub))
load("/data/bioinf/projects/data/2019_MAITcellRNAseq/de/withoutXLOC/RData/09c_res_c_CTRL_K12.RData")  # res_filtered
load("/data/bioinf/projects/data/2019_MAITcellRNAseq/de/withoutXLOC/RData/04_bmTable.RData")  # bmTable
new <- left_join(as_tibble(sub) %>% mutate("ensembl_gene_id"=rownames(sub)), bmTable)
#make names unique
new$external_gene_name <- sub('[.]', '_', make.names(new$external_gene_name, unique = TRUE))
rownames(new) <- new$external_gene_name
counts <- new[,1:10]
rownames(counts) <- rownames(new)

#counts <- counts[-grep("NA_", rownames(counts)), , drop = FALSE]

dds_K12 <- DESeqDataSetFromMatrix(countData = counts,
                              colData = sampleTable_K12_unstim,
                              design = design)
#getting rid of 0 count genes (some artifact from joining)
keep <- rowSums(counts(dds_K12)) > 0
dds_K12 <- dds_K12[keep,]
dds_K12
save (dds_K12, file= file_dds_K12)

```


Pre-filtering
by excluding low count genes we reduce the memory usage of the later methods
```{r}
keep <- rowSums(counts(dds_K12)) >= 10
dds_K12 <- dds_K12[keep,]
dds_K12
```


Set factor levels
It may be important to define in which order the conditions are compared. (default is alphabetical order)
not sure if this is needed here. Might return later
```{r}
#dds$condition <- factor(dds$condition, levels = c("T1","T2"))
```
The following steps are to visualize and cluster the data. For this process a transformed version of the data is useful.
rlog-transformation

```{r}
### - takes >1 min so only recompute if necessary - ###
recompute <- 0

file_vsd_K12 <- file.path(results_dir, "03_vsd_K12.Rdata")
file_rld_K12 <- file.path(results_dir, "04_rld_K12.Rdata")
if (recompute) {
  vsd_K12 <- varianceStabilizingTransformation(dds_K12, blind = FALSE)
  rld_K12 <- rlog(dds_K12, blind = FALSE)
  save(vsd_K12, file = file_vsd_K12)
  save(rld_K12, file = file_rld_K12)
} else {
  load(file = file_vsd_K12)
  load(file = file_rld_K12)
}
head(assay(rld_K12), 3)

```
Plot showing dependence of sd on mean
```{r}
#plot of sd and mean
ntd_K12 <- normTransform(dds_K12)
recompute <- 0
if(recompute){


  meanSdPlot(assay(rld_K12))
  jpeg(file = file.path(graphics_dir, "02_rld_meanSd_K12.jpeg"))
  meanSdPlot(assay(rld_K12))
  dev.off()
  meanSdPlot(assay(vsd_K12))
  jpeg(file = file.path(graphics_dir, "03_vsd_meanSd_K12.jpeg"))
  meanSdPlot(assay(vsd_K12))
  dev.off()

}
```
For K12 vsd has the most even mean!

Heatmap of the sample-to-sample distances





```{r}
sampleDists <- dist(t(assay(vsd_K12)))
sampleDistMatrix <- as.matrix(sampleDists)

#row.names(sampleTable) <- c("T_A", "C_A", "T_D", "C_D", "T_B", "C_B", "T_C", "C_C", "T_E", "C_E")
# annotation of rows
rows <- match(rownames(sampleDistMatrix), rownames(sampleTable))
annotation <- data.frame(treatment = sampleTable$treatment[rows])

#rownames(annotation) <- rownames(sampleDistMatrix)
row.names(sampleDistMatrix) <- c("T1", "C1", "T4", "C4", "T2", "C2", "T3", "C3", "T5", "C5")
colnames(sampleDistMatrix) <- c("T1", "C1", "T4", "C4", "T2", "C2", "T3", "C3", "T5", "C5")
rownames(annotation) <- rownames(sampleDistMatrix)
save(annotation, file = file.path(results_dir, "annotation"))
# heatmap colors
colors <- colorRampPalette( rev(brewer.pal(9, "Greys")) )(255)
# save heatmap in image file
file <- paste0(graphics_dir, "05_heatmap_s2s-distance_K12.pdf")
#jpeg("pheatmap.jpg")
pheatmap(sampleDistMatrix,
         color=colors,
         #border_color = NA,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         treeheight_row = 20,
         treeheight_col = 20,
         cellwidth=25,
         cellheight=25,
         #annotation_row = annotation,
         annotation_col = annotation,
         #annotation_colors = ann_colors,
         legend = F,
         show_colnames = T,
         show_rownames = T,
         #main=paste0("Sample-2-sample distance of (rlog-)transformed count data\n"),
         #filename=file
)

```


PCA
Shows samples in 2D plot, the axes representing their two principal components
```{r}

jpeg(file=file.path(graphics_dir, "05_vsd_K12_PCA.jpeg"))
plotPCA(vsd_K12)
dev.off()
```
Looks fine, no outliers
Clustering is completed.

```{r}
save(dds_K12, file = file.path(results_dir, "05_dds_pre_processed.Rdata"))
```
Standard differential expression analysis
This method goes through all steps of DESeq2 and gives results.
```{r}
recompute <- 1
# runtime
start <- Sys.time()

file_res_K12 <- file.path(results_dir, "06_res_K12.Rdata")
file_dds_K12 <- file.path(results_dir, "07_dds_K12.Rdata")
if (recompute) {
  dds_K12 <- DESeq(dds_K12)
  res_K12 <- results(dds_K12)
  save(dds_K12, file = file_dds_K12)
  save(res_K12, file = file_res_K12)
} else {
  load(file = file_dds_K12)
  load(file = file_res_K12)
}

res_K12
```

Perform LFC transformation
For better visualization and gene ranking we use the method lfcShrink
```{r}
resultsNames(dds_K12)
resLFC <- lfcShrink(dds_K12, coef = "condition_K12_vs_16h_Ctrl", type = "apeglm")
resLFC
```
Table ordered by p-value
For a better overview we can order the table by p-value and give a summary of how many p-values were <0.1
Finally, we can get a results table.
```{r}
resOrdered_K12 <- res_K12[order(res_K12$pvalue),]
resOrdered_K12
summary(res_K12)
#How many below 0.1?
sum(res_K12$padj < 0.1, na.rm = TRUE)
#results table
res05_K12 <- results(dds_K12, alpha = 0.05)
summary(res05_K12)
sum(res05_K12$padj < 0.05, na.rm = TRUE)
```


Export the list of genes that pass p-value

```{r}
resSig_K12 <- subset(resOrdered_K12, padj < 0.1)
resSig_K12
write.csv(as.data.frame(resSig_K12), file = file.path(results_dir, "08_results_K12_p01_no_exclusion"))
write.csv(as.data.frame(resOrdered_K12), file = file.path(results_dir, "09_results_all.csv"))
#runtime
runtime <- Sys.time() - start
capture.output(runtime, file = file.path(results_dir, "10_runtime.txt"))
```
